---
title: "week1_repdata"
output: html_document
---

##Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded here 
##Data Source
**Dataset** Available here: [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)


**Variables:**

* _steps_: Number of steps taking in a 5-minute interval (missing values are coded as NA)

* _date_: The date on which the measurement was taken in YYYY-MM-DD format

* _interval_: Identifier for the 5-minute interval in which measurement was taken

##Loading and preprocessing the data

```{r, echo=TRUE}
#load the activity data
activity <- read.csv('activity.csv', sep = ',', header = TRUE)
#clean up date
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?
```{r scatterplot, fig.height=4}
#Total number of steps taken each day
dailysteps <- aggregate(steps ~ date, data=activity, FUN="sum")
#ploat steps
plot(activity$date,activity$steps,ylab='Number of Steps', xlab="Date", type="l")

```
```{r, echo=TRUE}
#calculate mean and median of steps
step_mean <- mean(tapply(activity$steps,activity$date,sum), na.rm=TRUE)
step_median <- median(tapply(activity$steps,activity$date,sum), na.rm=TRUE)
```

```{r, echo=TRUE}
print(paste0("Mean total number of steps taken per day is: ", step_mean))

```

```{r, echo=TRUE}
print(paste0("Median total number of steps taken per day is: ", step_median))
```

##What is the average daily activity pattern
```{r, echo=TRUE}
 #steps by 5 mintue intervals
  dailysteps_interval <- aggregate(steps ~ interval, data=activity, FUN="sum")
  plot(dailysteps_interval$interval,dailysteps_interval$steps,ylab='Number of Steps', xlab="5 minute intervals", type="l")
  #find the interval with the max number of steps recorded
  max_interval <- dailysteps_interval$interval[which.max(dailysteps_interval$steps)]
  #plot the interval to the max!
  abline(v=(max_interval), lty=3, col="blue")
```

```{r, echo=TRUE}
print(paste0("The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is ...  ", max_interval))
```

##Imputing missing values

```{r, echo=TRUE}
#total number of rows with na's
total_na <- sum(is.na(activity))
```

```{r, echo=TRUE}
print(paste0("Total number of rows with na's is .... ", total_na))
```

```{r, echo=TRUE}
#To account for missing data, I will use new data with imputed values using the mean
activity_new <- merge(activity, dailysteps, by="date", suffixes=c("", ".mean"))
nas <- is.na(activity_new$steps)
activity_new$steps[nas] <- activity_new$steps.mean[nas]
activity_new <- activity_new[, c(1:3)]
```

```{r, echo=TRUE}
#plot new data
dailysteps_new <- aggregate(steps ~ date, data=activity_new, FUN="sum")
plot(activity_new$date,activity_new$steps,ylab='Number of Steps', xlab="", type="l")
```

```{r, echo=TRUE}
#calculate mean and median of steps with new data
step_mean <- round(mean(tapply(activity_new$steps,activity_new$date,sum), na.rm=TRUE))
step_median <- round(median(tapply(activity_new$steps,activity_new$date,sum), na.rm=TRUE))
```

It appears imputing data had no effect on the averages.
```{r, echo=TRUE}
print(paste0("Mean total number of steps taken per day with new data is: ", step_mean))

```

```{r, echo=TRUE}
print(paste0("Median total number of steps taken per day with new data is: ", step_median))
```

##Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
# functin to identify weekend or weekday
  daytype <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
      "weekend"
    } else {
      "weekday"
    }
  }
  # convert daytime from character to factor
  activity_new$daytype <- as.factor(sapply(activity_new$date, daytype))
  
  # aggreagate new data and include weekend or weekday value based on mean
  dailysteps_new <- aggregate(steps ~ interval + daytype, activity_new, mean)
  library(lattice)
  
  #2 plots showing average from weekend and weekday
  dayofweek <- xyplot(steps ~ interval | daytype, data=dailysteps_new, layout=c(2,1), type='l')
  #trellis.device(device="png", filename='figures/weekend_weekday.png')
  print(dayofweek)
```

Based on the plot the activity appears higher across the weekend than during the week. The weekday graph shows lower overall activity with the exception of mondays.
