---
title: "week1_repdata"
output: html_document
---

##Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded here 
##Data Source
**Dataset** Available here: [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)


**Variables:**

* _steps_: Number of steps taking in a 5-minute interval (missing values are coded as NA)

* _date_: The date on which the measurement was taken in YYYY-MM-DD format

* _interval_: Identifier for the 5-minute interval in which measurement was taken

##Loading and preprocessing the data

```{r, echo=TRUE}
#load the activity data
activity <- read.csv('activity.csv', sep = ',', header = TRUE)
#clean up date
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?
```{r scatterplot, fig.height=4}
#Total number of steps taken each day
dailysteps <- aggregate(steps ~ date, data=activity, FUN="sum")
#ploat steps
plot(activity$date,activity$steps,ylab='Number of Steps', xlab="Date", type="l")

```
```{r, echo=TRUE}
#calculate mean and median of steps
step_mean <- mean(tapply(activity$steps,activity$date,sum), na.rm=TRUE)
step_median <- median(tapply(activity$steps,activity$date,sum), na.rm=TRUE)
```

```{r, echo=TRUE}
print(paste0("Mean total number of steps taken per day is: ", step_mean))

```

```{r, echo=TRUE}
print(paste0("Median total number of steps taken per day is: ", step_median))
```

##What is the average dail activity pattern
```{r, echo=TRUE}
 #steps by 5 mintue intervals
  dailysteps_interval <- aggregate(steps ~ interval, data=activity, FUN="sum")
  plot(dailysteps_interval$interval,dailysteps_interval$steps,ylab='Number of Steps', xlab="5 minute intervals", type="l")
  #find the interval with the max number of steps recorded
  max_interval <- dailysteps_interval$interval[which.max(dailysteps_interval$steps)]
  #plot the interval to the max!
  abline(v=(max_interval), lty=3, col="blue")
```

```{r, echo=TRUE}
print(paste0("The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is ...  ", max_interval))
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
